---
title: "Generate Molecular Initiating Event Predictions for Selected MIEML Classifiers"
author: "Joseph Bundy"
date: "3/11/2025"
output: 
  html_notebook:
    code_folding: none
    toc: true
    toc_depth: 1
    toc_float: false
    number_sections: false
---

<br>

This notebook uses MIEML functions to generate MIE predictions from high performance classifiers

## Data Import

<br>

Define constants to be used throughout notebook 
```{r}
REFCHEMDB_PATH = "../data/refchemdb/NIHMS1537541-supplement-Supplement1.xlsx"
GENE_EXPRESSION_DATA_PATH = "../data/gene_expression/normalized_gene_expression_data.rds"
METADATA_PATH = "../data/mongodb_dump/httr_metadata.rds"
```
 
<br>
 
Import classifier performance summary and exemplar chemical table
```{r}
classifier_performance_summary <- read.csv(file = "../data/classifier_performance_summary.csv", header = TRUE)
exemplar_chems <- read.csv(file = "../data/exemplar_chems.csv", header = TRUE)
```

<br>

Re-derive exemplar reference chemical annotations
```{r}
#load mieml_httrpl functions
source("../scripts/ML_functions_HTTr.R")
collapsed_refchemdb_targets <- collapse_refchemdb_targets(refchemdb_path = REFCHEMDB_PATH,
                                                        support_levels = c(3),
                                                        min_chemicals = 5,
                                                        cutree_h = 0.7,
                                                        cluster_method= "complete")

refchem_collapsed <- collapsed_refchemdb_targets$collapsed_targets
refchem_collapsed <- refchem_collapsed[refchem_collapsed$support >= 3,]

#filter refchemdb targets such that each chemical is linked to only one target
stingy_target_annotations <- data.frame()
for (i in unique(refchem_collapsed$dsstox_substance_id)){
        relevant_targets <- refchem_collapsed[refchem_collapsed$dsstox_substance_id == i,]
        if(nrow(relevant_targets) > 0){
                relevant_targets <- relevant_targets[order(relevant_targets$support, decreasing = TRUE),]
                stingy_target_annotations <- rbind(head(relevant_targets, 1), stingy_target_annotations)
        }
}
refchem_collapsed <- stingy_target_annotations
rm(stingy_target_annotations)
```
 

## Generate MIE predictions

Use MIEML function to generate MIE predictions for classifiers that passed empirical significance analysis
```{r, warning= FALSE}
#subset the table of classifier performance metrics to only those that passed empirical significance analysis
high_performance_classifiers <- classifier_performance_summary[classifier_performance_summary$int_acc_emp_pval <= 0.1 & classifier_performance_summary$hold_acc_emp_pval <= 0.1,]

#generate MIE predictions
MIE_predictions <- generate_MIE_predictions(high_performance_models = high_performance_classifiers,
                                            exemplar_chems = exemplar_chems,
                                            metadata_path = METADATA_PATH,
                                            gene_expression_data_path = GENE_EXPRESSION_DATA_PATH,
                                            target_annotations = refchem_collapsed)

#save results
saveRDS(MIE_predictions, file = "../data/MIEML_predictions.rds")
```

Print session info
```{r}
sessionInfo()
```


